# HDCytoData


## Overview

This package contains a set of publicly available high-dimensional cytometry data sets, formatted into `SummarizedExperiment` and `flowSet` objects. The objects contain expression values as well as meta-data including sample IDs, group IDs (where relevant), population labels, and protein marker names.

These data sets have been used in our previous publications for benchmarking evaluations. They are provided here as `SummarizedExperiments` and `flowSets` to make them easier to access for other method developers.

The raw data sets are also available from FlowRepository (see help files for each data set).



## How to load data

The data sets are stored on Bioconductor's ExperimentHub.

```{r}
library("HDCytoData")
```


The data objects can either be loaded either by referring to the data set name, or by loading via ExperimentHub:

```{r}
# Load data object by name

# 'SummarizedExperiment' format
Bodenmiller_BCR_XL_SE()

# 'flowSet' format
Bodenmiller_BCR_XL_flowSet()
```


```{r}
# Load via ExperimentHub
eh <- ExperimentHub()

# 'SummarizedExperiment' format
query(eh, "Bodenmiller_BCR_XL_SE")
eh[["Bodenmiller_BCR_XL_SE"]]

# 'flowSet' format
query(eh, "Bodenmiller_BCR_XL_flowSet")
eh[["Bodenmiller_BCR_XL_flowSet"]]
```



## Details

Details about each data set are provided in the help files. For example:

```{r}
Bodenmiller_BCR_XL_SE
```


Code to generate the data objects from the raw data files is included in `inst/scripts/make-data.R`.


